---
title: "circlss among circular-response regression packages"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
# Static article: nothing here is executed at build time. Every fit is run once
# offline (dev/comparison/figs.R) and its console output pasted below the matching
# chunk; the figures are pre-generated into figures/. Keeps the build free of the
# comparator packages and of fitting that is not bit-stable across BLAS.
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

`circlss` provides capacity to fit **distributional regression for circular responses**: every parameter of a circular distribution gets its own penalized-spline predictor, REML-selected through `mgcv::gam()`, across a library of circular families.

## Where circlss sits

| Package | Approach | Inference | Families | Nonlinear effects | Per-parameter | Mixtures | Geometries |
|---|---|---|---|---|---|---|---|
| **circlss** | mgcv penalized-spline GAMLSS | Freq. (REML/EFS) | **12** | splines (+cyclic) | every param | EM | c~l, c~c, l~c |
| circular | `lm.circular` | Freq. ML | vM | linear | location | — | c~l, c~c |
| Directional | `spml.reg` | Freq. ML | proj. normal | linear | location (+scale) | — | c~l, c~c |
| bpnreg | `bpnr` | Bayes (MCMC) | proj. normal | linear | location (2D) | mixed eff. | c~l |
| brms | `von_mises` | Bayes (Stan) | vM | `s()`/`t2()` | κ | `mixture()` | c~l, c~c |
| NPCirc | kernel (NW/LL) | Nonparam. | vM kernel | kernel | mean only | — | c~l, c~c, l~c |
| BAMBI | `fit_angmix` | Bayes/ML | vM, wN, vmsin/cos | none | — | yes | — (density) |

```{r install}
# Every comparator is on CRAN:
install.packages(c("circular", "NPCirc", "bpnreg", "brms", "Directional", "BAMBI"))
```

## 1. `c~l` regression: circular response, linear covariate

Fisher & Lee's 31 periwinkles: direction moved (circular) on distance moved (linear) — the shared example of `circular` and `NPCirc`.

```{r peri-data}
library(NPCirc); library(circular)
data(periwinkles)
dist  <- periwinkles$distance
dir_c <- circular(periwinkles$direction, units = "degrees")            # for NPCirc
peri  <- data.frame(distance = dist, direction = periwinkles$direction * pi / 180)
```

**circular**: a von Mises link `mu = mu0 + 2*atan(b*x)`; location only, one global concentration.

```{r peri-circular}
lm.circular(y = circular(peri$direction), x = dist, type = "c-l", init = 0)
```

```
#>  Circular-Linear Regression
#>  Coefficients:    Estimate Std. Error t value Pr(>|t|)
#>             [1,] -0.008317   0.001359   6.119  4.7e-10 ***
#>  Summary: (mu in radians)   mu: 2.426 (0.112)   kappa: 3.224 (0.716)
```

**NPCirc**: a nonparametric conditional-mean curve (Nadaraya–Watson).

```{r peri-npcirc}
kern.reg.lin.circ(dist, dir_c, t = NULL, bw = 12.7, method = "NW")
```

```
#> Call: kern.reg.lin.circ(x = dist, y = dir_c, bw = 12.7, method = "NW")
#> Data: dist (31 obs.);  Bandwidth 'bw' = 12.7        Rho: 0.94
```

**circlss**: a von Mises GAMLSS.

```{r peri-circlss}
library(circlss)
b <- circ_gam(list(direction ~ s(distance, k = 5),   # mu(distance), penalized smooth
                            ~ distance),             # log-kappa LINEAR in distance
              data = peri, family = vmlss())
summary(b)
predict(b, newdata = data.frame(distance = quantile(dist, c(0, .5, 1))), type = "response")
```

```
#> Family: vmlss     Link function: tanhalf log
#>
#> Parametric coefficients:
#>               Estimate Std. Error z value Pr(>|z|)
#> (Intercept)    1.00525    0.08014  12.543  < 2e-16 ***   # mu intercept
#> (Intercept).1 -0.52074    0.56203  -0.927    0.354       # log-kappa intercept
#> distance.1     0.04798    0.01014   4.729 2.25e-06 ***   # log-kappa slope on distance
#>
#> Approximate significance of smooth terms:
#>             edf Ref.df Chi.sq p-value
#> s(distance)   1      1  20.64 6.4e-06 ***
#>
#> Deviance explained =   47%      -REML = 28.222     n = 31
#>
#>          mu    kappa
#> 0%    1.819    0.623     # near-uniform heading at the shortest distance
#> 50%   1.586    5.399
#> 100%  1.020  206.939     # near-deterministic at the longest
```

**brms**: von Mises smooth via Stan. Here `kappa ~ 1`: the `kappa ~ distance` term that circlss fits (κ → 207) overflows Stan on 31 points, so brms is held at constant κ.

```{r peri-brms}
library(brms)
pb <- transform(peri, direction = atan2(sin(direction), cos(direction)))   # (-pi, pi]
brm(bf(direction ~ s(distance), kappa ~ 1), data = pb, family = von_mises(),
    chains = 2, iter = 1000, seed = 1)
```

```
#>                 Estimate Est.Error   Q2.5 Q97.5
#> Intercept          1.323     0.207  0.971 1.797
#> kappa_Intercept    1.084     0.237  0.563 1.488
#> sdistance_1       -2.217     2.507 -7.464 2.828
```

![**periwinkles · c~l.** Cylinder (left), mean direction (centre), concentration κ on a log scale (right). All four fits agree on the mean. The right panel is what only circlss draws: κ climbing 0.6 → 207 with distance (log-κ slope *p* = 2×10⁻⁶), the ±1 circ-SD band (centre) pinching shut as the snails' headings tighten. `circular` holds κ flat; `NPCirc` has no κ; `brms`, overflowed at `kappa ~ distance`, is held at κ~1 — and there circlss(κ~1, dashed) sits on brms's mean curve: the same Bayesian smooth, by REML.](figures/cmp-1-periwinkles.png)

## 2. One distribution, three inference engines — projected normal

`bpnreg`'s Motor data: 42 hand-movement phases under three priming conditions. The projected normal writes an angle as `atan2(mu2, mu1)` of two linear predictors, the same form `bpnr` (Gibbs), `spml.reg` (ML), and circlss `pnlss` (REML) all fit.

```{r motor-data}
library(bpnreg); library(Directional)
data(Motor)        # Phaserad (circular) ~ Cond (3 levels) + AvAmp
```

```{r motor-fits}
fit_b <- bpnr(Phaserad ~ Cond + AvAmp, data = Motor, its = 5000, burn = 1000, seed = 1)
round(coef_lin(fit_b)[, c("mean", "sd")], 3)                  # bpnreg, Gibbs

X <- model.matrix(~ Cond + AvAmp, Motor)[, -1]
spml.reg(Motor$Phaserad, X, rads = TRUE)$be                  # Directional, ML

b <- circ_gam(list(Phaserad ~ Cond + AvAmp, ~ Cond + AvAmp),  # circlss pnlss, REML
              data = Motor, family = pnlss())
summary(b)
```

```
#> bpnreg coef_lin (component I rows 1-4, II rows 5-8)        Directional $be
#>                mean    sd                                         cos(y)  sin(y)
#> (Intercept)   1.337 0.452       (Intercept)                       1.338   1.401
#> Condsemi.imp -0.476 0.637       Condsemi.imp                     -0.382  -1.126
#> Condimp      -0.605 0.656       Condimp                          -0.522  -0.907
#> AvAmp        -0.010 0.012       AvAmp                            -0.009  -0.011
#> ...                                                   # circlss pnlss summary:
#> (Intercept).1 1.397 0.438       (Intercept).1  1.39719  z = 3.192  0.0014 **
#> Condsemi.imp.1 -1.162 ...        Condsemi.imp.1 -1.16182  z = -1.953  0.051 .
#> Deviance explained = 12.9%   -REML = 65.46   n = 42
```

![**Motor · c~l, projected normal.** At the reference condition the three engines trace one curve: the fitted direction (centre) and the resultant length R (right) coincide, their ±1 circ-SD bands overlapping — agreement on the whole distribution, not just the mean. circlss delivers bpnreg's posterior-mean fit and Directional's ML fit by penalized REML, no MCMC; and through mgcv, `Cond + AvAmp` could become `Cond + s(AvAmp)` — a smooth neither neighbour offers.](figures/cmp-2-motor.png)

## 3. `c~c` circular-circular regression

### wind direction, 6 a.m. to noon

```{r wind-cc}
data(wind)
i6 <- seq(7, 1752, by = 24); i12 <- seq(13, 1752, by = 24)   # daily 06:00, 12:00
wr <- function(a) atan2(sin(a), cos(a))
dcc <- data.frame(w6 = wr(wind$wind.dir[i6]), w12 = wr(wind$wind.dir[i12]))

kern.reg.circ.circ(circular(dcc$w6), circular(dcc$w12), bw = 6.1, method = "NW")
# winding response -> projected normal (a von Mises mean can't follow a heading
# that goes round the circle); mean AND concentration each a cyclic smooth:
b_cc <- circ_gam(list(w12 ~ s(w6, bs = "cc"), ~ s(w6, bs = "cc")),
                 data = dcc, family = pnlss())
summary(b_cc)$s.table
```

```
#> Family: pnlss          n = 73        Deviance explained = 46%
#>           edf Ref.df Chi.sq p-value
#> s(w6)   2.894      8  33.66  <2e-16 ***   # mean direction smooth
#> s.1(w6) 2.974      8  31.04  <2e-16 ***   # concentration smooth
```

![**wind · c~c, torus.** The noon heading winds around the torus as a function of the 6 a.m. heading. NPCirc traces the mean (kernel); circlss `pnlss` fits the mean (edf 2.9) **and** the concentration (edf 3.0), both *p* < 2×10⁻¹⁶ — the varying spread (right) is the part the kernel has no parameter for.](figures/cmp-3cc-wind.png)

## 4. `l~c` linear-circular regression

### does wind speed depend on direction?

```{r wind-lc}
data(speed.wind2)
sw <- na.omit(data.frame(dir = speed.wind2$Direction, speed = speed.wind2$Speed))
sw$rad <- wr(sw$dir * pi / 180)

kern.reg.circ.lin(circular(sw$dir, units = "degrees"), sw$speed, method = "LL")
# model BOTH the mean and the SD (gausslss, tau = 1/sigma) of speed by direction:
b_lc  <- circ_gam(list(speed ~ s(rad, bs = "cc"), ~ s(rad, bs = "cc")),
                  data = sw, family = gausslss())
# mean alone under GCV (the less-conservative selector a kernel resembles):
b_gcv <- circ_gam(speed ~ s(rad, bs = "cc"), family = gaussian(), method = "GCV.Cp", data = sw)
summary(b_lc)$s.table; summary(b_gcv)$s.table
```

```
#> gausslss            edf Ref.df Chi.sq p-value
#>   s(rad)   [mean]  0.001      8  0.001   0.465    # mean: flat under REML
#>   s.1(rad) [SD]    2.016      8  8.027   0.009    # SD: varies with direction
#> gaussian (GCV)      edf Ref.df     F p-value
#>   s(rad)   [mean]  6.436      8  1.815   0.028    # mean: wiggly under GCV
```

![**speed.wind2 · l~c, can.** The **mean** is selector-dependent: NPCirc's kernel and circlss-under-GCV (green/orange, centre) trace a weak undulation; circlss-under-REML draws it flat — a borderline effect on the bias–variance dial. The **scale** is not ambiguous (right): the SD swings 3.3 → 4.7 m/s with direction (edf 2.0, *p* = 0.009). The structure lives in the spread — which a mean-only kernel cannot see.](figures/cmp-3lc-speed.png)

## 5. Mixtures — BAMBI vs circ_mix

Not a regression: `BAMBI` fits fixed-form mixtures of toroidal densities. For the 8TIM protein's 490 (φ, ψ) backbone dihedrals, a four-component bivariate sine von Mises captures the Ramachandran basins. `circ_mix` reaches the same clusters by EM over circular GAMs, each component a conditional f(ψ | φ) × marginal f(φ).

```{r tim8}
library(BAMBI); data(tim8)
fit_angmix("vmsin", tim8, ncomp = 4, n.iter = 500, n.chains = 1)        # BAMBI, HMC

circ_mix(list(psi ~ cos(phi) + sin(phi), phi ~ 1),                      # circ_mix, EM
         data = tim8, family = vmlss(), K = 4)
```

```
#> BAMBI: 4 component vmsin mixture, 490 obs, fitted via HMC.
#>
#> Finite mixture of circular GAMs (circ_mix) -- joint torus density
#>   family vmlss | K = 4 components | 490 obs        BIC = 1751.77
#>   components (MAP):  pi = 0.293 / 0.474 / 0.200 / 0.032   (n = 152 / 236 / 86 / 16)
#>   converged in 95 iterations; restart basin hits 1/10.
```

![**tim8 · mixtures**, same (φ, ψ) coordinates, each panel coloured by its own MAP clusters (labels aligned by overlap, so a basin reads the same colour both sides). Both recover the four Ramachandran basins (β-sheet, α-helix, left-handed, bridge). `BAMBI` is Bayesian and covariate-free; `circ_mix` is built on `circ_gam`, so the components that here cluster a static density can instead carry the smooths and covariates of §1–3.](figures/cmp-5-mixture.png)